\documentclass[12pt,a4paper]{article}
\pagestyle{headings}

\usepackage[utf8x]{inputenc}
\usepackage{ucs}
\usepackage[british]{babel}
\usepackage[nodayofweek]{datetime}
\usepackage{hyperref}

\author{Michael Walker}
\title{Version Control and Teamwork}

\begin{document}

\maketitle
\tableofcontents
\pagebreak

\section{Version Control}

A version control system (VCS) is a tool (or collection of tools)
which make working on multiple versions of something, be it a program,
website, document, whatever, and keeping track of all of the changes
much easier. If it helps, you can think of a VCS as a glorified undo
tool.

If you're used to developing software by first making a backup of the
file you want to work on, and then editing the original file (or even
just editing the original file); or by editing the live system rather
than an offline copy; you need version control.

The work-flow for version control is a bit different:

\begin{description}
  \item[Edit] the appropriate files, there is no need to take a backup
    first.
  \item[Test] your changes. Of course, you did this before, right?
  \item[Commit] your changes, that is, tell the VCS which files you
    have changed and (usually) enter a short message explaining what
    you did.
  \item[Deploy] your commits, that is, send your commits to a server,
    or some other user. If you're the sole user of the VCS for your
    project, this may not be necessary (but can still be good---more
    on that later).
\end{description}

You may think that having to enter a message for each commit is a
hassle. Well, consider that by entering a good message, you'll be able
to see what the commit did. This may be incredibly useful down the
line when you're trying to find out where you introduced a bug, or
removed something that was needed. By using a VCS, you're not just
giving yourself an easy way to undo things, you're building a
detailed history of the development of your project, which is an
invaluable tool in itself.

Eventually, you will make a commit that you realise was a bad
idea. It's inevitable. It's ok. You can tell people you were drunk
when you did it. Fortunately, we come to another strength of VCSs
here, undoing things. Every VCS (that I know of) has some way to tell
the system to undo a specific commit. How this is implemented
varies---the commit may simply vanish, or the system may insert a new
commit reverting all of its changes. You may also need to do some
manual intervention, if you're undoing something which you have since
built upon.

Finally, let's cover branches and merging in this introduction. These
are key concepts in the VCS workflow, and entire development paradigms
have sprung up around how to use branches correctly (we'll be looking
at two). A branch is a copy of your code. Each branch can be distinct,
and branches can be merged (perhaps with some manual intervention
required) at any time as well. It's possibly easiest to explain with
a couple of examples of typical branches you may have:

\begin{description}
  \item[master] this is typically your stable, production-ready
    code.
  \item[devel] depending on your workflow, you may also have a
    development branch. This will be ahead of master, and will be
    merged into master when you have implemented everything you want
    for the next release
  \item[hotfixes] again, depending on your workflow, you may have a
    branch for emergency fixes to releases. This will also be ahead of
    master, but only by a handful of commits at a time, after which
    point it will be merged and a new release made.
\end{description}

Of course, where branches really come into their own is when you have
a good workflow based around them. It's also typical to have ``feature
branches'', one branch for each new feature you are implementing,
which is them merged into the development branch when complete.

\subsection{Distributed vs. Centralised}

There are two approaches when it comes to designing version control
systems, and which one your system uses will drastically affect how
you work with it. These approaches are distributed (or decentralised)
and centralised systems.

\begin{description}
  \item[Centralised] systems have one central server which houses the
    repository. If you commit something, it is sent to the server, if
    you lose access for a period of time you can't keep working and
    deploy your commits later. Despite this downside, because everyone
    is working off the same copy of the code, conflicts when merging
    may be reduced.
  \item[Distributed] systems do not have a central server. If you
    commit something, it is saved on your local machine. When you
    deploy your commits, that simply consists of you sending them to
    another user, who merges them into their copy. As there is no
    server to lose access to, you can keep working at all
    times. However, because everyone has their own version of the
    code, which may have many commits difference with someone else,
    merge conflicts may be more common.
\end{description}

In practice, distributed systems are much more convenient to use than
centralised systems, and the potential problem of very inconsistent
repositories is not often realised. Furthermore, users typically use a
hybrid of the two approaches, a distributed system with one
``canonical'' version of the code, which may be guarded by one
privileged user. There may even be multiple levels of this canonical
code, forming a tree structure with trusted users at each level
accepting things from the level below which pass the quality checks,
possibly doing some more work of their own, and then pushing their
code up the tree to be reviewed by the next person. This sort of
structure is often used in large open source projects, for example,
the Linux kernel. Linus Torvalds has control over the canonical kernel
code, and reviews all commits to it and either accepts or denies.

A typical workflow for distributed systems is to have an account on a
service like GitHub, which provides a globally-accessible repository
which you can push your code to. This allows you to work anywhere even
if you didn't bring your local copy of the code with you, provides
free backup of all of your commits, and enables people working with
you to frequently merge your commits and so reduce potential
conflicts.

\subsection{Some popular VCS tools}

There are a lot of version control tools. Below are some of the major
ones, which is it often convenient to have at least a basic knowledge
of, even if you use one of them exclusively for everything you do:

\begin{description}
  \item[Concurrent Versions System (CVS)] centralised, very old.
  \item[Subversion (svn)] centralised, intended to be a better CVS.
  \item[Git] distributed, has gained a lot of popularity in the open
    source community.
  \item[Mercurial (hg)] distributed, has gained some popularity in the
    open source community.
\end{description}

I find that centralised tools are usually used in organisations with a
strict hierarchy, and distributed tools in organisations where every
developer is (more or less) equal. If nothing else, an interesting
modern manifestation of how the Cathedral and the Bazaar methods
differ.

If you're going to join us in working on the DDNS project, we shall be
using Git and GitHub extensively.

\pagebreak
\section{Using Git}

This is just going to be a quick command reference, if you want more
details, you can read the manpages or look online.

\begin{description}
  \item[\texttt{git init}] Initialise a new git repository in the
    current directory.
  \item[\texttt{git add}] Tell git about a changed file or directory.
  \item[\texttt{git mv}] Move a file
  \item[\texttt{git rm}] Delete a file
  \item[\texttt{git commit}] Commit a set of changes
  \item[\texttt{git log}] Look at the git history
  \item[\texttt{git checkout}] Create and change branches
  \item[\texttt{git remote}] Manage remote repositories
    (\textit{e.g.} your github repo)
  \item[\texttt{git push}] Push commits to a remote
  \item[\texttt{git fetch}] Fetch (but do not merge) changes from a
    remote
  \item[\texttt{git merge}] Merge another branch (including things
    fetched from a remote).
  \item[\texttt{git tag}] Give a name (\textit{e.g.} a version
    number) to the latest commit.
\end{description}

Now, we're going to look at two workflows common with git, which you
use depends on both personal preference and how the project you are
working on is structured.

\subsection{Git Flow}

\href{http://nvie.com/posts/a-successful-git-branching-model/}{Git
  Flow} is a method of development with Git for projects that have a
clear release schedule, where one branch is to be used to contain the
production-ready code. It is a good methodology to use when working
with a lot of other people, or on some large or important system, as
it is very strict about how development and releases are done.

There are two main branches that always are present:

\begin{description}
  \item[master] the main branch, containing production-ready
    code. Every commit should be a new release.
  \item[develop] the main development branch, containing code for the
    next release.
\end{description}

\texttt{develop} is, of course, almost always ahead of
\texttt{master}. When the code in \texttt{develop} reaches a point
suitable for a new release, it is merged into \texttt{master}.

Git Flow uses a number of support branches: feature branches, release
branches, and hotfix branches. Each one can only be branched off from
certain branches and can only merge back into certain other
branches. This rigid structure helps to ensure development progresses
smoothly.

\subsubsection{Feature branches}

These can only branch off from \texttt{develop} and can only be merged
back into \texttt{develop}.

These are used to develop features for some future release, although
which release may be unknown. The idea is that once a feature is done,
it will be merged into \texttt{develop} when it is desired to go into
the next release (it will wait until that time) or, if the feature
doesn't work out, the branch will just be deleted.

A feature branch is created by branching from \texttt{develop}:

\begin{verbatim}
$ git checkout -b FEATURE develop
\end{verbatim}

And, when finished, will be merged back into \texttt{develop}:

\begin{verbatim}
$ git checkout develop
$ git merge --no-ff FEATURE
$ git branch -d FEATURE
$ git push origin develop
\end{verbatim}

The flag \texttt{--no-ff} causes the merge to be stored as a new
commit, even when this is unnecessary. The reason for this is that it
is easier to see in the history when a feature branch starts and when
it stops, which makes removing a feature (if necessary) much easier.

\subsubsection{Release branches}

These can only branch off from \texttt{develop} and must merge back
into both \texttt{develop} and \texttt{master}.

These are used to prepare a new release. When the code in
\texttt{develop} is in a suitable state, a release branch is created
and the first commit to the branch is updating the version
number. After that, there may be some number of commits to finalise
the code for release. Once it has been finalised, it is merged into
\texttt{master} (creating a new release) and \texttt{develop}.

A release branch is created by branching from \texttt{develop}:

\begin{verbatim}
$ git checkout -b release-VERSION develop
update the version number
$ git commit -a -m "Updated the version number to VERSION"
\end{verbatim}

And, when finished, will be merged into both \texttt{master} and
\texttt{develop}:

\begin{verbatim}
$ git checkout master
$ git merge --no-ff release-VERSION
$ git tag -a VERSION
$ git checkout develop
$ git merge --no-ff release-VERSION
$ git branch -d release-VERSION
\end{verbatim}

\subsubsection{Hotfix branches}

These can only branch off from \texttt{master} and must merge back
into both \texttt{develop} and \texttt{master}.

A hotfix branch is like a release branch, however it is used to create
an emergency new release to fix an urgent bug in the production code
that cannot wait until the next release.

A release branch is created by branching from \texttt{master}:

\begin{verbatim}
$ git checkout -b hotfix-VERSION master
update the version number
$ git commit -a -m "Updated the version number to VERSION"
\end{verbatim}

Then, some commits are made to fix the bug, and it is merged into both
\texttt{master} and \texttt{develop}:

\begin{verbatim}
$ git checkout master
$ git merge --no-ff hotfix-VERSION
$ git tag -a VERSION
$ git checkout develop
$ git merge --no-ff hotfix-VERSION
$ git branch -d hotfix-VERSION
\end{verbatim}

\subsection{Github Flow}

\section{Some Best Practices for Teamwork}

\end{document}